---
title: "Umatilla River Hydrology Pipeline — Working Notebook"
author: "temperheic"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: '3'
    df_print: paged
  html_notebook:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: flatly
subtitle: '01 → 01b → 01c → 01d → 02: Acquisition, Regulation Correction, Record Extension,
  Stream Power'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  warning    = FALSE,
  message    = TRUE,
  fig.width  = 10,
  fig.height = 6,
  fig.retina = 2,
  comment    = "#>"
)

library(tidyverse)
library(lubridate)
library(knitr)
library(scales)

# Base data directory — adjust if needed
DATA_DIR <- "data/"
```

# Overview

This notebook documents the hydrology pipeline for the Umatilla River
discharge–channel migration analysis (DOGAMI O-25-10 follow-up). The pipeline
runs through five scripts in sequence:

| Script | Purpose | Key Output |
|--------|---------|------------|
| `01_hydrology_acquisition.R` | Pull USGS daily + peak flows for 3 gages | `peak_flows.csv`, `daily_flows.csv`, flood frequency |
| `01b_mckay_creek_acquisition.R` | Pull McKay Creek data from USGS + USBR Hydromet | MCKO, MYKO, MCK daily records |
| `01c_regulation_correction.R` | Reconstruct unregulated peaks at 14033500 | `reconstructed_peaks.csv` |
| `01d_move3_extension_and_weighting.R` | MOVE.3 extend 14020850 using corrected index | `move3_extended_record.csv` |
| `02_reach_attributes_and_scaling.R` | CMZ import, DA scaling, stream power | `reach_attributes.csv` |

The sequencing is intentional: 01c must run before 01d because the MOVE.3
extension of 14020850 uses the *corrected* 14033500 as its primary index gage.

---

# 1. Gage Network and Data Acquisition (01)

## 1.1 Gage metadata

Three USGS gages span the study area from the Blue Mountains headwaters to the
Columbia Plateau:

```{r gage-metadata, eval=TRUE}
gage_metadata <- tribble(
  ~gage_id,    ~gage_name,                                       ~position,      ~DA_sqmi, ~record,
  "14020000",  "Umatilla R above Meacham Cr nr Gibbon",          "upstream",     131,      "1933–2024",
  "14020850",  "Umatilla R at W Reservation Bndy nr Pendleton",  "mid",          682,      "1996–2025",
  "14033500",  "Umatilla R near Umatilla",                       "downstream",   2290,     "1904–2024"
)

kable(gage_metadata, caption = "Umatilla River gage network")
```

Reach assignment follows the gage positions:

- **RS 30–39** (upstream of Pendleton) → 14020000 (Gibbon, 131 mi²)
- **RS 9–29** (Pendleton corridor) → 14020850 (Reservation Bndy, 682 mi²)
- **RS 1–8** (lower river) → 14033500 (near Umatilla, 2290 mi²)

## 1.2 Data retrieval

`01_hydrology_acquisition.R` uses `dataRetrieval` to pull daily mean discharge,
annual instantaneous peaks, and site metadata from USGS NWIS. Each retrieval
function (`pull_daily_flow()`, `pull_peak_flow()`, `pull_site_info()`) is atomic
— one gage, one data type — then mapped across gages with `purrr::map()` and
`safely()` for fault tolerance.

```{r peak-flows-load, eval=TRUE}
# Load peak flows from the pipeline output
peak_flows <- read_csv(file.path(DATA_DIR, "peak_flows.csv"),
                       col_types = cols(.default = col_guess(),
                                        gage_id = col_character()))
```

```{r peak-summary, eval=TRUE}
peak_flows %>%
  group_by(gage_id) %>%
  summarize(
    n_peaks    = n(),
    first_year = min(water_year),
    last_year  = max(water_year),
    median_q   = median(peak_q_cfs, na.rm = TRUE),
    max_q      = max(peak_q_cfs, na.rm = TRUE),
    .groups    = "drop"
  ) %>%
  kable(caption = "Peak flow record summary", digits = 0,
        format.args = list(big.mark = ","))
```


## 1.3 Initial flood frequency (L-moments)

The initial flood frequency uses L-moment LP3 fitting — adequate for bankfull
(Q2) estimation and preliminary stream power context. The full B17C analysis
(EMA, regional skew, MGBT) runs through `peakfq` in the scripts that need it.

```{r flood-freq-initial}
# Hardcoded from pipeline run — L-moment LP3
ff_initial <- tribble(
  ~gage_id,    ~Q2,   ~Q5,    ~Q10,    ~Q25,    ~Q50,    ~Q100,
  "14020000",  2119,  3361,   4212,    5247,    6054,    6880,
  "14020850",  5644,  9451,  12399,   16470,   19776,   23331,
  "14033500",  5293,  8747,  11332,   14822,   17612,   20547
)

ff_initial %>%
  mutate(across(Q2:Q100, ~ comma(.x))) %>%
  kable(caption = "Initial flood frequency estimates (L-moment LP3, cfs)")
```

**Key issue:** All 121 peaks at 14033500 carry NWIS qualification codes 5 or 6
(regulation-affected) due to McKay Dam. The L-moment fit doesn't account for
this. The EMA/B17C analysis with regional skew (SIR 2016-5083) gives slightly
different numbers:

```{r flood-freq-ema}
# Hardcoded from peakfq::emafit() run
ff_ema <- tribble(
  ~gage_id,    ~Q2,   ~Q10,    ~Q100,  ~record_length, ~weighted_skew,
  "14020000",  2119,   4212,    8010,   92,              0.389,
  "14020850",  5542,  12199,   23909,   30,              0.120,
  "14033500",  5223,  11144,   21899,  121,              0.245
)

ff_ema %>%
  mutate(across(Q2:Q100, ~ comma(.x))) %>%
  kable(caption = "B17C flood frequency estimates (EMA, weighted skew, cfs)")
```

Regional skew from SIR 2016-5083: G~r~ = −0.07, SE = 0.424.

---

# 2. McKay Creek Data Acquisition (01b)

## 2.1 Why McKay Creek matters

McKay Dam (USBR, completed 1927) regulates McKay Creek, a major tributary
entering the Umatilla at RS 26. Pre-dam, McKay Creek peaked at 1,460–3,250
cfs/yr with an average daily flow of ~160 cfs. Post-dam, peaks >1,500 cfs have
occurred only 3 times since 1927. The dam has 65,534 AF active storage plus a
~6,000 AF flood pool.

This regulation suppresses peaks at 14033500 (downstream of the confluence),
which means the entire 121-year record at that gage reflects an artificial flow
regime. To characterize natural flood frequency for RS 1–8, we need to
reconstruct what the peaks would have been without the dam.

## 2.2 Data sources

`01b_mckay_creek_acquisition.R` pulls from two sources:

1. **USGS NWIS** — gage 14023500 (McKay Cr nr Pendleton): daily + peaks,
   ~1918–1991. Includes 5 pre-dam peaks (WY 1919–1923) that represent natural
   hydrology.

2. **USBR Hydromet** — via `https://www.usbr.gov/pn-bin/daily.pl`:
   - **MCKO**: McKay Cr below dam, daily Q (cfs), 1980–present
   - **MYKO**: McKay Cr above reservoir (natural inflow), daily Q (cfs), 1980–present
   - **MCK**: McKay Reservoir storage, daily (AF), 1980–present

Cross-validation of MCKO vs USGS 14023500 during their 1980–1991 overlap showed
r = 1.0 and mean difference = 0 cfs — MCKO is a seamless extension.

## 2.3 The 2020 flood as proof of concept

```{r mckay-2020, eval=FALSE}
# This chunk is for reference — the actual data are in the 01b outputs
# WY 2020 event: MYKO natural inflow peaked at 2,470 cfs
#                MCKO dam release peaked at 373 cfs (one month later)
#                Attenuation: 85%
# This confirms McKay Dam is a powerful peak regulator.
```

The WY 2020 flood illustrates the problem: McKay Creek's natural inflow peak
was 2,470 cfs (MYKO), but the dam released only 373 cfs (MCKO) — a peak one
month later, at 85% attenuation. Any flood frequency analysis using the
observed 14033500 record captures this attenuated signal, systematically
underestimating the natural flood regime.

---

# 3. Regulation Correction (01d)

## 3.1 Approach

The regulation correction reconstructs what each annual peak at 14033500 would
have been if McKay Dam didn't exist. The "regulation deficit" is defined as:

$$\text{deficit} = Q_{\text{natural inflow (MYKO)}} - Q_{\text{dam release (MCKO)}}$$

This deficit is added back to the observed Umatilla peak:

$$Q_{\text{unregulated}} = Q_{\text{observed}} + \text{deficit}$$

For the **truth period** (1980–present), MYKO data exist and the deficit is
computed directly. For the **pre-MYKO post-dam period** (1927–1979), the
natural inflow must be estimated. Two methods were tested:

- **Method (i): Statistical scaling** — log-log regression of MYKO peak vs
  Umatilla peak with optional seasonal harmonics and Duan's smearing correction.
- **Method (ii): Daily mass balance** — back-calculate reservoir inflow from
  storage changes and dam releases: Q~in~ = dS/dt + Q~out~.

## 3.2 Event pairing

For each of the 121 annual peaks at 14033500, a ±14-day window around the peak
date is used to extract concurrent McKay Creek conditions (MYKO, MCKO, MCK
storage). This is handled by `pair_peaks_with_mckay()`.

## 3.3 Cross-validation

Both methods were validated against the truth period (45 events with concurrent
MYKO + MCKO data):

```{r cv-performance}
cv_perf <- tribble(
  ~method,                ~level,    ~n,  ~rmse,  ~bias,    ~pct_bias, ~nse,    ~r_squared,
  "statistical_scaling",  "deficit", 45,  630,    -18,      -2.6,       0.251,   0.374,
  "mass_balance",         "deficit", 45,  367,    -148,    -17.3,       0.783,   0.837,
  "statistical_scaling",  "inflow",  45,  451,     14,       2.6,       0.418,   0.520,
  "mass_balance",         "inflow",  45,  274,    -120,    -13.5,       0.785,   0.856
)

cv_perf %>%
  mutate(across(c(rmse, bias), ~ round(.x)),
         across(c(pct_bias, nse, r_squared), ~ round(.x, 3))) %>%
  kable(caption = "Cross-validation performance: regulation deficit and inflow estimation")
```

**Mass balance was decisively better:** NSE = 0.783 vs 0.251 at the deficit
level. It has a −17% bias (daily storage smooths within-day peaks) but much
lower variance than statistical scaling (R² = 0.837 vs 0.374). Mass balance was
auto-selected for the reconstruction.

For 1927–1979 peaks where MCK storage data are unavailable, the statistical
model provides the fallback.

## 3.4 Correction summary

```{r correction-summary}
correction_summary <- tribble(
  ~period,              ~correction_source,    ~n_peaks, ~mean_obs, ~mean_unreg, ~mean_deficit, ~max_deficit, ~mean_pct,
  "pre_dam",            "none_predam",         23,       4704,      4704,         0,             0,            0.0,
  "post_dam_no_myko",   "statistical_scaling", 52,       5590,      6620,         1030,          3280,        13.8,
  "post_dam_with_myko", "observed_myko",       45,       6780,      7880,         1100,          3340,        12.5
)

correction_summary %>%
  mutate(across(mean_obs:max_deficit, ~ comma(.x))) %>%
  kable(caption = "Correction summary by period")
```

Key findings:

- **94 of 97 post-dam peaks** were corrected upward (deficit > 0)
- Average regulation deficit: ~1,000–1,100 cfs (~13–14% of peak)
- Maximum deficit: ~3,300 cfs for large events
- WY 1993 had a negative deficit (irrigation release exceeded natural inflow)
- Pre-dam peaks (1904–1926) are left unchanged

---

# 4. MOVE.3 Record Extension (01c)

## 4.1 The problem

USGS 14020850 (Pendleton) has only a 30-year record (1996–2025). The 2020 flood
(largest on record by a wide margin) sits in this short sample, inflating the
upper tail of the frequency curve. MOVE.3 (Bulletin 17C Appendix 8) extends the
record using longer-record index gages, stabilizing the frequency estimates.

## 4.2 Why use the corrected 14033500?

The original 01c used the **regulated** 14033500 record as one index gage. The
regulation signal from McKay Dam adds noise to the log-space correlation with
14020850 — the index gage's peaks are systematically low relative to the natural
hydrologic relationship. Using the corrected series (from 01c) removes this
artifact and yields a tighter regression.

This is the key reason 01c runs before 01d: the corrected peaks feed directly
into the MOVE.3 regression.

## 4.3 Concurrent period diagnostics

```{r concurrent-diagnostics}
concurrent_diag <- tribble(
  ~index_gage,                  ~version,    ~n_concurrent, ~r_log,
  "14033500 (regulated)",       "Old 01c",   29,            0.9291,
  "14033500 (corrected)",       "New 01c",   29,            0.9550,
  "14020000 (Gibbon)",          "Both",      29,            0.9182
)

concurrent_diag %>%
  kable(caption = "Log-space Pearson correlation with 14020850 during concurrent period (1996–2024)",
        digits = 4)
```

The correlation improved from **0.929 to 0.955** after correction — a meaningful
tightening in a 29-year concurrent window. Removing the regulation noise at the
index gage cleaned up the relationship.

## 4.4 MOVE.3 regression parameters

```{r move3-diagnostics}
move3_diag <- tribble(
  ~index,                    ~version,  ~n1, ~ne, ~r,      ~b_move, ~adj_mean, ~adj_sd,
  "14033500 (regulated)",    "Old",     29,  26,  0.9291,  1.0256,  3.7692,    0.2732,
  "14033500 (corrected)",    "New",     29,  26,  0.9550,  1.0426,  3.7632,    0.2702,
  "14020000 (Gibbon)",       "Both",    29,  26,  0.9182,  1.0667,  3.7173,    0.2745
)

move3_diag %>%
  kable(caption = "MOVE.3 regression diagnostics (log-space)",
        digits = 4)
```

The corrected index has b~MOVE~ = 1.0426 (vs 1.0256 regulated). The corrected
index peaks have slightly more variance in log-space (the deficit restoration
spread the peaks out), pulling the slope closer to a 1:1 relationship.

## 4.5 Flood frequency comparison

```{r frequency-comparison}
ff_comparison <- tribble(
  ~method,                            ~Q2,   ~Q5,   ~Q10,   ~Q25,   ~Q50,   ~Q100,
  "Original (30 yr)",                 5542,  9263,  12199,  16448,  20007,  23909,
  "MOVE.3 via 14033500 (regulated)",  5900,  9644,  12443,  16303,  19395,  22662,
  "MOVE.3 via 14033500 (corrected)",  5792,  9415,  12145,  15938,  19001,  22260,
  "MOVE.3 via 14020000 (Gibbon)",     5149,  8404,  10937,  14566,  17583,  20870,
  "Weighted MOVE.3 (old, regulated)", 5514,  9020,  11692,  15443,  18501,  21780,
  "Weighted MOVE.3 (new, corrected)", 5468,  8915,  11553,  15271,  18314,  21588
)

ff_comparison %>%
  mutate(across(Q2:Q100, ~ comma(.x))) %>%
  kable(caption = "Flood frequency comparison at 14020850 (cfs)")
```

```{r pct-change}
# Compute % change from original
orig <- ff_comparison %>% filter(method == "Original (30 yr)")

ff_comparison %>%
  mutate(across(Q2:Q100, ~ round(100 * (.x - orig[[cur_column()]]) /
                                   orig[[cur_column()]], 1))) %>%
  kable(caption = "Percent change from original 30-year record",
        digits = 1)
```

The corrected-index MOVE.3 converges toward the Gibbon estimates across the
board (1.5–2 percentage points closer than the regulated version). The weighted
combination now gives Q100 = 21,588 cfs (vs 21,780 with the old regulated
index) — the two index gages tell a more internally consistent story.

## 4.6 Diagnostic plots

```{r move3-plots, eval=TRUE, fig.width=8, fig.height=5, out.width="100%"}
# These plots execute if the PNGs exist from the 01c pipeline run
plot_files <- c(
  "move3_regression_corrected.png",
  "move3_regression_gibbon.png",
  "move3_frequency_comparison.png"
)

for (f in plot_files) {
  fpath <- file.path(DATA_DIR, f)
  if (file.exists(fpath)) {
    cat(paste0("\n\n### ", tools::file_path_sans_ext(f), "\n\n"))
    knitr::include_graphics(fpath)
  }
}
```


---

# 5. Reach Attributes and Stream Power (02)

## 5.1 CMZ data import

`02_reach_attributes_and_scaling.R` imports the DOGAMI CMZ summary spreadsheets
(Umatilla River and McKay Creek) and builds a unified reach table with geometry,
erosion rates, avulsion history, confinement classification, and discharge
estimates.

```{r reach-table, eval=TRUE}
# Load reach table if available
reach_path <- file.path(DATA_DIR, "reach_attributes.rds")
if (file.exists(reach_path)) {
  reach_tbl <- read_rds(reach_path)

  reach_tbl %>%
    select(rs, rs_num, assigned_gage, avg_width_ft, slope, sinuosity,
           median_rate_cw, confinement, q2_reach_cfs, omega_wm2) %>%
    kable(caption = "Reach attributes summary (selected columns)",
          digits = c(0, 0, 0, 0, 4, 2, 4, 0, 0, 1))
} else {
  message("reach_attributes.rds not found — run 02 first")
}
```

## 5.2 Drainage area scaling

Discharge at each reach is estimated via power-law scaling from the assigned
gage:

$$Q_{\text{reach}} = Q_{\text{gage}} \times \left(\frac{DA_{\text{reach}}}{DA_{\text{gage}}}\right)^\alpha$$

where α is estimated from the log-log regression of Q2 vs DA across the three
gages.

## 5.3 Unit stream power

Unit stream power (ω, W/m²) is the primary forcing metric:

$$\omega = \frac{\gamma \, Q \, S}{w}$$

where γ = 9810 N/m³, Q is the reference discharge (Q2 as bankfull proxy),
S is channel slope, and w is bankfull width.

---

# 6. Conceptual Framework — What Gets Used Where

The pipeline produces multiple discharge products that serve different purposes
in the downstream analysis:

```{r usage-table}
usage <- tribble(
  ~reaches,   ~purpose,                          ~discharge_product,
  "RS 30–39", "Flood frequency + stream power",  "14020000 observed peaks (unregulated)",
  "RS 9–29",  "Flood frequency + stream power",  "14020850 MOVE.3 extended (weighted, from 01d)",
  "RS 1–8",   "Natural flood frequency",          "14033500 corrected peaks (from 01c)",
  "RS 1–8",   "Migration modeling (actual regime)","14033500 observed (regulated) peaks"
)

kable(usage, caption = "Discharge product assignment by reach group")
```

The dual use at RS 1–8 reflects a conceptual distinction:

- **Natural flood frequency** (corrected record): answers "what magnitude floods
  *should* occur at this location absent human modification?" — needed for
  comparison with geomorphic thresholds and for climate projection baselines.

- **Migration modeling** (regulated record): answers "what flow regime actually
  *moved* the channel?" — the historical migration data postdate McKay Dam, so
  the empirical discharge–migration relationship should use the regulated flows
  that actually drove the observed channel changes.

---

# 7. Next Steps

With the hydrology pipeline complete, the path forward is:

1. **Run B17C** on the weighted MOVE.3 extended record (14020850) and on the
   corrected 14033500 record via `peakfq_app()`. This gives defensible frequency
   curves at all three gage positions.

2. **Finalize 02** — update to consume CSV outputs from the refactored pipeline;
   compute stream power at multiple return periods (not just Q2).

3. **Phase 1a** — develop the empirical discharge–migration model linking flood
   magnitude to observed erosion rates and avulsion events.

4. **Phase 2** — apply the model to projected future hydrology (RMJOC-II
   climate scenarios) to assess how changing flow regimes may alter geomorphic
   work.

---

# Session Info

```{r session-info}
sessionInfo()
```
